One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-06-30 doc/genesetdp.tex
Updates after comments from reviewer and ES.

\documentclass[a4paper,american]{lipics-v2016}
%\usepackage[USenglish]{babel}
%\usepackage{microtype}
\bibliographystyle{plainurl}
\title{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Null models for Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Acces}
%\EventNoEds{2}
\EventLongTitle{18th Workshop on Algorithms in Bioinformatics (WABI 2018)}
\EventShortTitle{WABI 2018}
\EventAcronym{WABI}
\EventYear{2018}
\EventDate{20–22 August 2018}
\EventLocation{Helsinki, Finland}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{equation}
p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
\label{eq:pval}
\end{equation}
where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.

We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
\begin{equation}
s=\sum_{i=1}^Q l_i.
\label{eq:sum_i}
\end{equation}
Such number-of-links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.

@@

We first implemented a random sampling algorithm to assess $N(s)$. We randomly selected sets of $Q$ genes from $\mathcal{G}$. These sets scores were calculated with Equation \ref{eq:sum_i}. We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets. We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}. We refer to this procedure as GeneSetMC.

@@

We also realized that one could calculate the full score distribution, $N(s)$, using dynamic programming. First, we define a function, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having $a$ links to the investigated pathway, and we define $R=\max_{i}{l_i}$. Furthermore, we define the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway. This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
\begin{equation}
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
\end{equation}
where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception of $N_{-1}(0,0)=1$.

@@

\begin{lstlisting}[caption={
The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.
}
, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1))
    N[0,0] = 1

@@

\[
N_a(s,c)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c)=D_a+N_{a-1}(s,c).
\]

This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements. That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ to previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$. For details see Listing \ref{lst:gensetdp}. We refer to this method as GeneSetDP.

@@

We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation \ref{eq:sum_i}, using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

@@

>>>>

\documentclass[a4paper,american]{lipics-v2016}
%\usepackage[USenglish]{babel}
%\usepackage{microtype}

\bibliographystyle{plainurl}

\title{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Null models for Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Acces}
%\EventNoEds{2}
%\EventShortTitle{WABI 2018}
%\EventAcronym{WABI}
%\EventYear{2018}
%\EventDate{20–22 August 2018}
%\EventLocation{Helsinki, Finland}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{equation}
p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
\label{eq:pval}
\end{equation}
where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation~(\ref{eq:sum_ij}).

We can reformulate Equation~(\ref{eq:sum_ij}) by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
\begin{equation}
s=\sum_{i=1}^Q l_i.
\label{eq:sum_i}
\end{equation}
Such number-of-links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.

@@

We first implemented a random sampling algorithm to assess $N(s)$. We randomly selected sets of $Q$ genes from $\mathcal{G}$. These sets scores were calculated with Equation~(\ref{eq:sum_i}). We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets. We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~(\ref{eq:pval}). We refer to this procedure as GeneSetMC.

@@

\begin{equation}
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
\end{equation}
where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception of $N_{-1}(0,0)=1$.

@@

\begin{lstlisting}[language=Python, caption={ The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.}, label=lst:gensetdp, captionpos=t, float, abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1))
    N[0,0] = 1

@@

\[
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c).
\]

This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements. That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add $\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b)$ to previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$. For details see Listing \ref{lst:gensetdp}. We refer to this method as GeneSetDP.

@@

We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation~(\ref{eq:sum_i}), using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation~(\ref{eq:pval}). We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

@@

To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC. As can be seen in Figure~\ref{fig:pscatter}, the relative difference is larger for lower $p$~values, likely due to the sampling errors of GeneSetMC.

\begin{figure}[htb]
  \begin{center}
		  \begin{tabular}[t]{c}
        FIXME: Make this scatterplot in log-log, GeneSetDP on x-axis, GeneSetMC on y-axis.
        (Idea from reviewer comment in email 2018-06-08.)
%				\includegraphics[width=1.0\textwidth]{FIXME}
		\end{tabular}
  \end{center}
  \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. We subsequently plotted the obtained $p$~values against each other .}
  \label{fig:pscatter}
\end{figure}
